---
title: "COVID-19 vaccine strategy simulations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
devtools::load_all()
library(ggplot2)
```

## to add

* example with vaccine rates different per group  
* example with antivaxxers  
* example with n arrivals per day with P(covid) (per age)

## Model Description

some equations...

## Vaccine effectiveness in Australia

Some key modelling parameters:  

* Contact matrix
* Age distribution  
* Imported cases (e.g. open border)  
* Vaccination strategy
* Vaccine effectiveness

### Demographics

For more details on the contact matrices, age distributions and their use see ....

```{r}
cm_oz <- import_contact_matrix("Australia","general")
nJ <- ncol(cm_oz)
dist_oz <- import_age_distribution("Australia")
SSv <- dist_oz*5e6
plot(dist_oz)
```

The 

```{r}
plot(cm_oz)
```

### Vaccination strategies

Even within the simplified population our model considers there are many possible vaccine allocation strategies. We explore a few below before considering how to incorporate them into the SEIR models. 

#### Random vaccination

One possible strategy would be to randomly vaccinate members of the population until. We might assume that this will lead to a vaccination rate per age group that is proportional to their percentage of the population. 

```{r}
random_vaccination <- function(n,p,s) {
    # n: number of available vaccines
    # p: probability of group j getting vaccinates
    # s: number of people in group j
    
    # number vaccinated
    nvac0 <- n*p
    
    # check if the population is fully vaccinated
    if (all(abs(ceiling(p)*s) < 0.01)) {
        nvac1 <- rep(0,length(p))
        return(nvac1)
    } else if (any((nvac0 - s) > 0.01)) {
    # check if any groups are (nearly) fully vaccinated
    # of so proportionally redistribute their vaccines
        overassigned <- nvac0 > s
        remaining_vac <- sum(nvac0 - s)
        nvac0 <- pmin(p*n,s*ceiling(p))
        p1 <- (p*overassigned)/sum(p*overassigned)
        nvac1 <- nvac0 + p1*remaining_vac
        return(nvac1)
    } else {
        return(nvac0)
    }
}
```

The graph below shows the pattern if we vaccinate 10,000 per day for population similar to Sydney.

```{r}
# simulate random vaccinate 
TT <- floor(365*2)
SSvt <- matrix(nrow = TT+1,ncol=length(SSv))
SSvt[1,] <- SSv
nvac_per_day <- 10000
for (t in 2:(TT+1)) {
    nvac_t <- random_vaccination(nvac_per_day,dist_oz,SSvt[(t-1),])
    SSvt[t,] <- SSvt[(t-1),] - nvac_t
}

# plot the resulting vaccination rate graph
SSvt <- data.frame(SSvt)
names(SSvt) <- seq(5,80,5)
SSvt$day <- 0:TT
SSvt <- tidyr::pivot_longer(SSvt,cols=-17,names_to="age_group",values_to="n")
SSvt$age_group <- as.numeric(SSvt$age_group)
ggplot(SSvt,aes(x=day,col=age_group,,group=factor(age_group),y=n/SSvt)) +
    geom_line(size=1) + 
    theme_bw(base_size = 16) +
    labs(x = "Time (days)", y= "Unvaccinated (n)") +
    scale_color_continuous(name="Age")
```

#### Random vaccination with decreasing efficacy

The random vaccination graph seems too linear. We might expect that the rate of vaccinations varies. For instance it might start off a little slow, then reach a peach before again slowing down as those too busy or hesitant to get the vaccine remain. For the more pessimistic case we may have somoething like the graph below.

```{r}
mm <- function(p,half) {
    # function that
    # p: proportion vaccinated
    # half: point at which vaccination rate halfs
    (p/(p+half))*(1.0+half)
}

df <- data.frame(p = seq(0,1,0.01),vac_rate = mm(seq(0,1,0.01),0.1))
ggplot(df,aes(x=p,y=vac_rate)) +
    geom_line() +
    geom_point() +
    coord_cartesian(ylim=c(0,1)) + 
    theme_bw()
```

How does this alter the linear curves above? First alter the allocation function.

```{r}
random_vaccination_mm <- function(n,p,s,s0,half) {
    # n: number of available vaccines
    # p: probability of group j getting vaccinates
    # s: number of people in group j
    
    # number vaccinated
    nvac0 <- n*p*mm(s/s0,half)
    
    # check if the population is fully vaccinated
    if (all(abs(ceiling(p)*s) < 0.01)) {
        nvac1 <- rep(0,length(p))
        return(nvac1)
    } else if (any((nvac0 - s) > 0.01)) {
    # check if any groups are (nearly) fully vaccinated
    # of so proportionally redistribute their vaccines
        overassigned <- nvac0 > s
        remaining_vac <- sum(nvac0 - s)
        nvac0 <- pmin(p*n,s*ceiling(p))
        p1 <- (p*overassigned)/sum(p*overassigned)
        nvac1 <- nvac0 + p1*remaining_vac
        return(nvac1)
    } else {
        return(nvac0)
    }
}
```

Then the results.

```{r}
# simulate random vaccinate 
TT <- floor(365*2)
SSvt <- matrix(nrow = TT+1,ncol=length(SSv))
SSvt[1,] <- SSv
nvac_per_day <- 10000
half <- 0.05
for (t in 2:(TT+1)) {
    nvac_t <- random_vaccination_mm(nvac_per_day,dist_oz,
                                    SSvt[(t-1),],SSv,half)
    SSvt[t,] <- SSvt[(t-1),] - nvac_t
}

# plot the resulting vaccination rate graph
SSvt <- data.frame(SSvt)
names(SSvt) <- seq(5,80,5)
SSvt$day <- 0:TT
SSvt <- tidyr::pivot_longer(SSvt,cols=-17,names_to="age_group",values_to="n")
SSvt$age_group <- as.numeric(SSvt$age_group)
ggplot(SSvt,aes(x=day,col=age_group,,group=factor(age_group),y=n)) +
    geom_line(size=1) + 
    theme_bw(base_size = 16) +
    labs(x = "Time (days)", y= "Unvaccinated (n)") +
    scale_color_continuous(name="Age")
```

#### COVID-19 case fatalility rate prioritised vaccination

We could vaccinate with a priority (probability) determined by the inverse of the case fatality rate (CFR) - which appears to be the Australia government plan. 

```{r}

```

#### Interaction with healthcare system prioritiesed vaccinations

An similar strategy to CFR strategy would be to vaccinate everyone who has an encounter with the healthcare system. Given issues with vaccine transport this would need to be hospitalisations or outpatient appointments (rather than GPs). We can get this data from ... 

#### Workplace and school prioritised vaccination 

An alternative strategy focused on limiting spread be the vaccination of people while at work, school or university. 

```{r}


```

### Vaccine effectiveness  

Two parameters control this. The first $VE_1$..

```{r echo=FALSE,fig.cap="*Figure 1:* Vaccination effectiveness",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph vac_eff1 {
  graph [overlap = false,
        fontsize = 10
        rankdir = LR]
  node [shape = box,
        fontname = Helvetica,
        color='grey']
  S[label='S\nsusceptible'];
  V[label='V\nvaccinated'];
  edge [color = 'red' style=dashed]
  S -> V
}",height=50)
```

The second $VE_2$ ...

```{r echo=FALSE,fig.cap="*Figure 2:* Vaccination effectiveness",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph vac_eff1 {
  graph [overlap = false,
        fontsize = 10
        rankdir = LR]
  node [shape = box,
        fontname = Helvetica,
        color='grey']
  V[label='V\nvaccinated'];
  Ev[label='Ev\nLatent period\n(vaccinated)'];
  edge [color = 'red' style=dashed]
  V -> Ev
}",height=50)
```


## Example simulation I - random vaccination

We'll use the following setup:  

* A population of 5 million  
* Vaccinate 20,000 people per day **randomly**. Nobody vaccinated at baseline. 
* Once we reach 60% vaccination the borders open. As a result there are imported cases. The number of arrivals in follows a $\text{Poisson}(\lambda)$ distribution where $\lambda = 200.0$ with a probability of COVID-19 of $p_{covid} = 0.05$ for all groups. The number of arrivals in each age group is assumed proportional to the age distribution of Australia. 
* Assume COVID has been extinguished in local populations, i.e. no local transmission in Australia or low and controlled enough to be ignorable in these average based models.
* Vaccine effectiveness of $VE_1 = 0.6$ and $VE_2 = 0.99$.

```{r}
# initial states
baseline <- 0.0
vaceff1 <- 0.99
vaceff2 <- 0.90
nvac <- 1000
S <- SSv*(1-baseline)
Sv <- SSv*baseline
E <- rep(0,nJ)
Ev <- rep(0,nJ)
I <- Iv <- rep(0,nJ)
R <- Rv <- rep(0,nJ)
state0 <- seir_cv_state0(S = S,E = E,I = I,R = R,Sv = Sv,Ev = Ev,Iv = Iv,Rv = Rv)

# functions
n_imp_cases <- function(t,vac_p,p) {
    rpois(100)*p[j]
}
nvac <- function(t) {
    # number of available vaccinations (total)
    1000*(t < 30) + 10000*(t >= 30)
}
param1 <- seir_cv_param(R0 = 2.5,
                        sigma=0.1,
                        gamma = 0.1,
                        cm=cm_oz,
                        dist=dist_oz,
                        vaceff1=vaceff1,
                        vaceff2=vaceff2,
                        nvac=nvac,
                        vac_alloc=,
                        n_imp=n_imp)

# simulation
res1 <- simulate_seir_cv(t = 1500,state_t0 = state0,param = param1)

# plot the results
plot(res1,y = c("E","Ev"))
plot(res1,y = c("Sv","S"))
plot(res1,y="incidence")
```

## Example simulation I - vaccine hesistancy

Consider the case where a certain section of the population will refuse to be vaccinated regardless of the consequences (e.g. no international travel). We can model this by increasing the heterogeneity of the population to account for the presence of a certain percentage of vaccine refusers in each age group.  

## ghg

0 imported cases.

```{r eval=FALSE, include=FALSE}
# parameters
baseline <- seq(0,1,0.2)
TT <- 1460
vaceff <- seq(0.8,1.0,0.1)

# results
df <- data.frame(t = numeric(length=0L),
                 E = numeric(length=0L),
                 base_group=character(length=0L),
                 vef_group=character(length=0L))

for (j in 1:length(vaceff)) {
    cat(j," of ",length(vaceff),"\n")
    for (i in 1:length(baseline)) {
        S <- SSv*(1-baseline[i])
        Sv <- SSv*baseline[i]
        E <- rep(10,nJ)
        Ev <- rep(0,nJ)
        I <- Iv <- rep(0,nJ)
        R <- Rv <- rep(0,nJ)
        state0 <- seir_cv_state0(S = S,E = E,I = I,R = R,Sv = Sv,Ev = Ev,Iv = Iv,Rv = Rv)
        param1 <- seir_cv_param(R0 = 2.5,
                                sigma=0.1,
                                gamma = 0.1,
                                cm=cm_oz,
                                dist=dist_oz,
                                vaceff=vaceff[j],
                                n_imp=0.0)
        res1 <- simulate_seir_cv(t = TT,state_t0 = state0,param = param1)
        df <- rbind(df,data.frame(t = 1:TT, 
                                  E = res1$epi["E"] + res1$epi["Ev"],
                                  base_group=paste0(100*baseline[i],"%"),
                                  vef_group=paste0(100*vaceff[j],"%")))
    }
}
```

```{r eval=FALSE, include=FALSE}
df$base_group <- factor(df$base_group,levels=paste0(100*baseline,"%"),ordered=TRUE)
df$vef_group <- factor(df$vef_group,levels=paste0(100*vaceff,"%"),ordered=TRUE)

p <- ggplot(df) +
    geom_line(aes(x = t, y = E, col = base_group)) +
    labs(x="Time (days)",y="Infections",size=1.0) +
    scale_color_discrete(name="Percentage\nvaccinated") +
    facet_wrap(~vef_group,ncol=1) +
    coord_cartesian(xlim=c(0,365*1.5)) +
    scale_y_continuous(breaks=seq(0,4e05,by=1e05),
        labels=format(seq(0,4e05,by=1e05),scientific =FALSE,big.mark = ",")) +
    theme_bw(base_size=14) +
    theme()
p
```

## Imported cases

10 imported cases per day.

```{r eval=FALSE, include=FALSE}
# parameters
# as above
# results
df1 <- data.frame(t = numeric(length=0L),
                 E = numeric(length=0L),
                 base_group=character(length=0L),
                 vef_group=character(length=0L))

for (j in 1:length(vaceff)) {
    cat(j," of ",length(vaceff),"\n")
    for (i in 1:length(baseline)) {
        S <- SSv*(1-baseline[i])
        Sv <- SSv*baseline[i]
        E <- rep(10,nJ)
        Ev <- rep(0,nJ)
        I <- Iv <- rep(0,nJ)
        R <- Rv <- rep(0,nJ)
        state0 <- seir_cv_state0(S = S,E = E,I = I,R = R,Sv = Sv,Ev = Ev,Iv = Iv,Rv = Rv)
        param1 <- seir_cv_param(R0 = 2.5,sigma=0.1,
                                gamma = 0.1,
                                cm=cm_oz,dist=dist_oz,
                                vaceff=vaceff[j],
                                n_imp=10)
        res1 <- simulate_seir_cv(t = T,state_t0 = state0,param = param1)
        df1 <- rbind(df1,data.frame(t = 1:T, 
                                  E = res1$epi["E"] + res1$epi["Ev"],
                                  base_group=paste0(100*baseline[i],"%"),
                                  vef_group=paste0(100*vaceff[j],"%")))
    }
}
```

```{r eval=FALSE, include=FALSE}
df1$base_group <- factor(df1$base_group,levels=paste0(100*baseline,"%"),ordered=TRUE)
df1$vef_group <- factor(df1$vef_group,levels=paste0(100*vaceff,"%"),ordered=TRUE)

p <- ggplot(df1) +
    geom_line(aes(x = t, y = E, col = base_group)) +
    labs(x="Time (days)",y="Infections",size=1.0) +
    scale_color_discrete(name="Percentage\nvaccinated") +
    facet_wrap(~vef_group,ncol=1) +
    coord_cartesian(xlim=c(0,365*1.5)) +
    scale_y_continuous(breaks=seq(0,4e05,by=1e05),
       labels=format(seq(0,4e05,by=1e05),scientific =FALSE,big.mark = ",")) +
    theme_bw(base_size=14) +
    theme()
p
```
