---
title: "Model Summary Report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(shiny)
library(ggplot2)
library(formattable)
library(sparkline)

### Re-run models and prepare graphs
## Could this be done more elegantly with a loop? surely?
## for(i in 1:6){I don't know how to loop through sequenced variables in R}

# Initialise an empty vector for initial states and initial parameters
# These get added to if the results have been saved in the report
summaryVec <- vector()
stateVec <- vector()
paramVec <- vector()
nameVec <- c(" ")

if(input$inclPlot1){
   mod_df_p1 <- prepData(df_epi=count$model_p1$epi, popsize=count$popSize_p1 , day1 = count$dateRange_p1[1])
   plot_p1 <- plotResults(df = mod_df_p1, scale=count$scale_p1, logScale=count$logScale_p1, plotvars=count$plotvars_p1, ndays=count$ndays_p1,
              xtraC=count$xtraC_p1, xtraP=count$xtraP_p1)
   summaryVec <- cbind(summaryVec, `Scenario 1` = summary <- summariseResults(count$model_p1$epi, 
                            count$state0_p1, 
                            count$param_p1, 
                            count$dateRange_p1, 
                            ig_t = count$intGeneral_t_p1,
                            ig_c = count$intGeneral_c_p1,
                            is_c = count$intSchool_c_p1,
                            iw_c = count$intWork_c_p1,
                            ih_c = count$intHome_c_p1
                            ))
   nameVec <- cbind(nameVec, "Scenario 1")
   stateVec <- cbind(stateVec, `Scenario 1` = formatC(round(count$state0_p1, digits=0), format="d", big.mark = ","))
   paramVec <- cbind(paramVec, `Scenario 1` = round(unlist(count$param_p1), digits=3))
}

if(input$inclPlot2){
   mod_df_p2 <- prepData(df_epi=count$model_p2$epi, popsize=count$popSize_p2 , day1 = count$dateRange_p2[1])
   plot_p2 <- plotResults(df = mod_df_p2, scale=count$scale_p2, logScale=count$logScale_p2, plotvars=count$plotvars_p2, ndays=count$ndays_p2, xtraC=count$xtraC_p2, xtraP=count$xtraP_p2)
      summaryVec <- cbind(summaryVec, `Scenario 2` = summary <- summariseResults(count$model_p2$epi, 
                            count$state0_p2, 
                            count$param_p2, 
                            count$dateRange_p2, 
                            ig_t = count$intGeneral_t_p2,
                            ig_c = count$intGeneral_c_p2,
                            is_c = count$intSchool_c_p2,
                            iw_c = count$intWork_c_p2,
                            ih_c = count$intHome_c_p2
                            ))
   nameVec <- cbind(nameVec, "Scenario 2")
   stateVec <- cbind(stateVec, `Scenario 2` = formatC(round(count$state0_p2, digits=0), format="d", big.mark = ","))
   paramVec <- cbind(paramVec, `Scenario 2` = round(unlist(count$param_p2), digits=3))
  }

if(input$inclPlot3){
   mod_df_p3 <- prepData(df_epi=count$model_p3$epi, popsize=count$popSize_p3 , day1 = count$dateRange_p3[1])
   plot_p3 <- plotResults(df = mod_df_p3, scale=count$scale_p3, logScale=count$logScale_p3, plotvars=count$plotvars_p3, ndays=count$ndays_p3, xtraC=count$xtraC_p3, xtraP=count$xtraP_p3)
      summaryVec <- cbind(summaryVec, `Scenario 3` = summary <- summariseResults(count$model_p3$epi, 
                            count$state0_p3, 
                            count$param_p3, 
                            count$dateRange_p3, 
                            ig_t = count$intGeneral_t_p3,
                            ig_c = count$intGeneral_c_p3,
                            is_c = count$intSchool_c_p3,
                            iw_c = count$intWork_c_p3,
                            ih_c = count$intHome_c_p3
                            ))
   nameVec <- cbind(nameVec, "Scenario 3")
   stateVec <- cbind(stateVec, `Scenario 3` = formatC(round(count$state0_p3, digits=0), format="d", big.mark = ","))
   paramVec <- cbind(paramVec, `Scenario 3` = round(unlist(count$param_p3), digits=3))
}

if(input$inclPlot4){
   mod_df_p4 <- prepData(df_epi=count$model_p4$epi, popsize=count$popSize_p4 , day1 = count$dateRange_p4[1])
   plot_p4 <- plotResults(df = mod_df_p4, scale=count$scale_p4, logScale=count$logScale_p4, plotvars=count$plotvars_p4, ndays=count$ndays_p4, xtraC=count$xtraC_p4, xtraP=count$xtraP_p4)
      summaryVec <- cbind(summaryVec, `Scenario 4` = summary <- summariseResults(count$model_p4$epi, 
                            count$state0_p4, 
                            count$param_p4, 
                            count$dateRange_p4, 
                            ig_t = count$intGeneral_t_p4,
                            ig_c = count$intGeneral_c_p4,
                            is_c = count$intSchool_c_p4,
                            iw_c = count$intWork_c_p4,
                            ih_c = count$intHome_c_p4
                            ))
   nameVec <- cbind(nameVec, "Scenario 4")
   stateVec <- cbind(stateVec, `Scenario 4` = formatC(round(count$state0_p4, digits=0), format="d", big.mark = ","))
   paramVec <- cbind(paramVec, `Scenario 4` = round(unlist(count$param_p4), digits=3))
}

if(input$inclPlot5){
   mod_df_p5 <- prepData(df_epi=count$model_p5$epi, popsize=count$popSize_p5, day1 = count$dateRange_p5[1])
   plot_p5 <- plotResults(df = mod_df_p5, scale=count$scale_p5, logScale=count$logScale_p5, plotvars=count$plotvars_p5, ndays=count$ndays_p5, xtraC=count$xtraC_p5, xtraP=count$xtraP_p5)
      summaryVec <- cbind(summaryVec, `Scenario 5` = summary <- summariseResults(count$model_p5$epi, 
                            count$state0_p5, 
                            count$param_p5, 
                            count$dateRange_p5, 
                            ig_t = count$intGeneral_t_p5,
                            ig_c = count$intGeneral_c_p5,
                            is_c = count$intSchool_c_p5,
                            iw_c = count$intWork_c_p5,
                            ih_c = count$intHome_c_p5
                            ))
   nameVec <- cbind(nameVec, "Scenario 5")
   stateVec <- cbind(stateVec, `Scenario 5` = formatC(round(count$state0_p5, digits=0), format="d", big.mark = ","))
   paramVec <- cbind(paramVec, `Scenario 5` = round(unlist(count$param_p5), digits=3))
}

if(input$inclPlot6){
   mod_df_p6 <- prepData(df_epi=count$model_p6$epi, popsize=count$popSize_p6, day1 = count$dateRange_p6[1])
   plot_p6 <- plotResults(df = mod_df_p5, scale=count$scale_p5, logScale=count$logScale_p5, plotvars=count$plotvars_p6, ndays=count$ndays_p6, xtraC=count$xtraC_p6, xtraP=count$xtraP_p6)
      summaryVec <- cbind(summaryVec, `Scenario 6` = summary <- summariseResults(count$model_p6$epi, 
                            count$state0_p6, 
                            count$param_p6, 
                            count$dateRange_p6, 
                            ig_t = count$intGeneral_t_p6,
                            ig_c = count$intGeneral_c_p6,
                            is_c = count$intSchool_c_p6,
                            iw_c = count$intWork_c_p6,
                            ih_c = count$intHome_c_p6
                            ))
   nameVec <- cbind(nameVec, "Scenario 6")
   stateVec <- cbind(stateVec, `Scenario 6` = formatC(round(count$state0_p6, digits=0), format="d", big.mark = ","))
   paramVec <- cbind(paramVec, `Scenario 6` = round(unlist(count$param_p6), digits=3))
 }

```


## Overview of simulations
```{r}

summary <- data.frame(
  cbind(
    c(
      "Simulation start date",
      "Simulation end date",
      "Simulation length (days)", 
      "Number susceptible at start of simulation", 
      "Number infectious at start of simulation", 
      "Number recovered at start of simulation", 
      "Assumed R0 at outbreak of epidemic",
      "Assumed duration of infection",
      "Intervention: Relative probability of transmission", 
      "Intervention: Relative contact rate (General)", 
      "Intervention: Relative contact rate (School)", 
      "Intervention: Relative contact rate (Work)", 
      "Intervention: Relative contact rate (Home)",
      "Peak of epidemic", 
      "Number infectious at peak" 
    ),
    summaryVec
  ), stringsAsFactors = FALSE
)

names(summary) <- nameVec

summary %>% 
  formattable(align = c("l", rep("r", NCOL(summary) - 1))) %>% 
  as.htmlwidget() %>% 
  spk_add_deps() 

```


## Simulation results

### `r if(input$inclPlot1){input$p1title}`
```{r}
if(input$inclPlot1){
plot_p1
}
```
`r if(input$inclPlot1){input$p1comment}`


### `r if(input$inclPlot2){input$p2title}`
```{r}
if(input$inclPlot2){
plot_p2
}
```
`r if(input$inclPlot2){input$p2comment}`

### `r if(input$inclPlot3){input$p3title}`
```{r}
if(input$inclPlot3){
plot_p3
}
```
`r if(input$inclPlot3){input$p3comment}`

### `r if(input$inclPlot4){input$p4title}`
```{r}
if(input$inclPlot4){
plot_p4
}
```
`r if(input$inclPlot4){input$p4comment}`

### `r if(input$inclPlot5){input$p5title}`
```{r}
if(input$inclPlot5){
plot_p5
}
```
`r if(input$inclPlot5){input$p5comment}`

### `r if(input$inclPlot6){input$p5title}`
```{r}
if(input$inclPlot6){
plot_p6
}
```
`r if(input$inclPlot6){input$p6comment}`




## Appendices

### Appendix 1. The SEIR Compartment Model

```{r}

    nodes <- data.frame(id = 1:4,

        # add labels on nodes
        label = c('S', 'E', 'I', 'R'),

        # size adding value
        size = c(rep(10, 4)),

        # Hierarchical level
        level = rep(0,4),

        # Group
        group = c("Susceptible", "Exposed", "Infectious", "Recovered"),

        # control shape of nodes
        shape = c(rep("square", 4)),

        font.color = c(rep("grey", 4)),

        # Don't need physics
        physics = rep(FALSE, 4),

        # tooltip (html or character), when the mouse is above
        title = c(HTML(paste0("S", br(), 'Susceptible')),
                  HTML(paste0("E", br(), "Exposed")),
                  HTML(paste0("I", br(), "Infectious")),
                  HTML(paste0("R", br(), "Recovered"))
        )

    )

    edges <- data.frame(from = seq(1,3),
        to = seq(2,4),
        id = LETTERS[seq(1,3)],
        dashes = c(rep(FALSE,3)),

        # Hover titles
        title = c(HTML('&beta;I/N'),
                  HTML('&sigma;'),
                  HTML('&gamma;')
                  )
    )


    visNetwork(nodes, edges) %>%
        visHierarchicalLayout(direction = "UD", levelSeparation = 80) %>%
        visEdges(arrows = "to") %>%
        visGroups(groupname = "Susceptible", color = "lightblue") %>%
        visGroups(groupname = "Susceptible", color = "orange") %>%
        visGroups(groupname = "Infectious", color = "red") %>%
        visGroups(groupname = "Recovered", color = "lightgreen") %>%
        visLegend(width = 0.1, position = "left", main = "Compartment", stepX = 50, stepY = 50) %>%
        visInteraction(hover = TRUE) %>%
        visPhysics(stabilization = FALSE) %>% 
        visEvents(startStabilizing = "function() {
            this.moveTo({scale:1.5})}") 

```

#### Summary of compartments
```{r}

initTable <- data.frame(
  Compartment = c("$S$",
                  "$E$",
                  "$I$",
                  "$R$"),

  Description = c("Susceptible",
                  "Exposed",
                  "Infectious",
                  "Recovered")
)

knitr::kable(initTable, escape = FALSE, caption = "Initial conditions for the model compartments")  %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = T, border_right = T, width = '20%') %>%
  column_spec(2, width = "80%")

```


### Appendix 2. Initial conditions

```{r}

initTable <- cbind(
  Compartment = c(rep("$S$",16),
                  rep("$E$",16),
                  rep("$I$",16),
                  rep("$R$",16)),

  Description = c(rep("Susceptible",16),
                  rep("Exposed",16),
                  rep("Infectious",16),
                  rep("Recovered",16)),

  `Age group` = paste0(seq(from=5, to=80, by=5), "-", seq(from=10, to=85, by=5)),

   stateVec
)

knitr::kable(initTable, escape = FALSE, caption = "Initial conditions for the model compartments")  %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = T, border_right = T)

```
