---
title: "Age structured epidemic models in COVOID"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{age-structure-mixing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE
)
```

```{r setup}
library(covoid)
library(ggplot2)
```

## Outline

This vignette introduces the COVOID programming API for modelling age-structured epidemic models in COVOID. It covers:

* The use of contact matrices.
* Modelling interventions/behaviour changes that reduce contact rates in different settings
* Modelling variation in infectiousness/susceptibility

## Base case - no interventions/behaviour changes

For our examples we'll consider an epidemic in a population of 5 million, modelled on the population of Sydney, Australia. We can import ... and plot... 

```{r,fig.width=5,fig.height=3,fig.retina=3,fig.align='center'}
# import contact matrix and age distribution
cm_oz <- import_contact_matrix("Australia","general")
p_age_oz <- import_age_distribution("Australia")
# use covoid plotting methods
plot(cm_oz)
```

```{r,fig.width=6,fig.height=3,fig.retina=3}
plot(p_age_oz)
```

These plots are ggplot objects (i.e. `class(plot(p_age_oz))` gives `r class(plot(p_age_oz))`) allowing further customisation beyond the basic default appearance. 

We'll start the epidemic simulation with ~50 cases, distributed geometrically between the ages of 20 and 80, roughly matching age distribution and early March incidence information available from [NSW Health](https://www.health.nsw.gov.au/Infectious/covid-19/Documents/covid-19-surveillance-report-20200430.pdf).

```{r}
# initial conditions
S <- p_age_oz*5e6
E <- c(0,0,0,0,ceiling(7*(0.9)^(1:12)))
I <- rep(0,length(S))
R <- rep(0,length(S))
```

We first consider the case without any intervention, letting the epidemic run its course. The epidemic peaks xx days after time zero, with an . If we are using a model with the code * we pass the parameters for the ODE system to the `*_param` function, the initial compartment states to `*_state0` and finally run the system using `simulate_*`. A `plot` method (built using [ggplot2](https://ggplot2.tidyverse.org/)) for the resulting object of class covoid allows quick visualisation of the evolution of the compartments over time.

```{r, fig.width=7,fig.height=3.5,fig.retina=3}
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm_oz,dist=p_age_oz)
state0 <- seir_c_state0(S = S,E = E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="No intervention")
```

## Interventions and behaviour change

We can add in time varying changes in person to person contact rates using the `contact_intervention` function. This function allows specification of the start and stop times of the intervention, along with potential delays until full adoption. These "interventions" are not only government mandated behaviour changes but any shift in inidividual behviour that could reasonably be expected to alter the levels of person-person physical interaction in the population under study. The below example reduces all contact by 20%. This could be considered to be simulating mild physical distancing. Note the ~50 day right shift in the epidemic trajectory. As we end the physical distancing before complete suppresion it still has the same outcome.

```{r, fig.width=7, fig.height=3.5,fig.retina=3}
# interventions
phys_dist <- contact_intervention(start = 10,stop = 150,reduce = 0.8,
                                 start_delay = 5,stop_delay = 5)
# model and simulation
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm_oz,dist=p_age_oz,
                     contact_intervention = phys_dist)
state0 <- seir_c_state0(S = S,E = E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="Physical distancing")
```

The COVOID package allows you to separate the contact rates into different settings - e.g. school, work and home. For example, using the "general" and "school" contact matrices for Australia we have the below patterns of interaction.

```{r,fig.height=2.5,fig.width=7,fig.retina=3}
cm_oz_all <- import_contact_matrix("Australia","general")
cm_oz_sch <- import_contact_matrix("Australia","school")
# separate out school and general population contact rates
cm_oz_all <- cm_oz_all - cm_oz_sch
p_all <- plot(cm_oz_all) + labs(title = "General") +
  theme(axis.text.x = element_text(size=6, angle=0),
          axis.text.y = element_text(size=6))
p_sch <- plot(cm_oz_sch) + labs(title = "School") +
  theme(axis.text.x = element_text(size=6, angle=0),
          axis.text.y = element_text(size=6))
gridExtra::grid.arrange(p_all,p_sch,ncol=2)
```

We can then add in time varying interventions using `create_intervention` to each setting. We do this by creating two `list` objects, one for the contact matrices and the other for the interventions. Each list must be named with the setting names matching across the contact matrix and intervention lists. For example, `all` and `sch` in the example below. For this example we again reduce general contact rates in the population by 20% to simulate physical distancing and combine it with a reduction in school age contact by 80% to simulate school closures. Again we see a slight ~25 day right shift in the epidemic trajectory with early ending of interventions leading to a similar final outcome to the no behaviour change setting.

```{r, fig.width=7, fig.height=3.5,fig.retina=3}
# contact matrices and interventions
cm <- list(all = cm_oz_all, sch = cm_oz_sch)
int <- list(sch=contact_intervention(start = 10,stop = 150,reduce = 0.2,
                                    start_delay = 5,stop_delay = 5),
           all=contact_intervention(start = 10,stop = 150,reduce = 0.8,
                                    start_delay = 5,stop_delay = 5))
# model and simulation
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm,
                      dist=p_age_oz,contact_intervention = int)
state0 <- seir_c_state0(S = S,E =E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="Physical distancing and school closures")
```

## Variation in infectiousness

The `im` argument to the `simulate_*_c` function allows us to vary infectiousness and susceptibility across age groups. An entry $p_{ij}$ of the transmissibility matrix $im$ denotes the **relative** chance of an individual in group $i$ infecting an individual in group $j$. For example if the average transmission probability is $p$ the entering $p_{ij} = 0.8$ indicates an 20% lower change of someone in group $i$ infecting someone in gorup $j$ compared to the average. In the example below we model those under 15 as being less infectious (but as as susceptible). Further using the `transmission_intervention` we allow for the adoption of general transmission reduction behaviours (e.g. everyone wearing facemasks/hand washing) across the population.

```{r, fig.width=7,fig.height=3.5,fig.retina=3}
# relative transmissibility matrix
im <- matrix(1,ncol=16,nrow=16)
im[,1:3] <- 0.8
# contact matrices and interventions
cm <- list(all = cm_oz_all, sch = cm_oz_sch)
int_c <- list(sch=contact_intervention(start = 10,stop = 150,reduce = 0.2,
                                      start_delay = 5,stop_delay = 5),
           all=contact_intervention(start = 10,stop = 150,reduce = 0.8,
                                    start_delay = 5,stop_delay = 5))
# transmissibility interventions
int_t <- transmission_intervention(start = 10,stop = 200,reduce = 0.9,
                                  start_delay = 5,stop_delay = 5)
# model and simulation
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm,dist=p_age_oz,
                     contact_intervention = int_c,
                     transmission_intervention = int_t,im = im)
state0 <- seir_c_state0(S = S,E =E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="Physical distancing; school
     closures;\nhandwashing/facemasks; and children less infectious")
```
