---
title: "Age structured epidemic models in covoid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{age-structure-mixing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(covoid)
library(ggplot2)
```

An epidemic in a population of 4 million. Starts uniformly across all age groups. Use Australian contact matrices and age distributions.

```{r}
cm_oz <- import_contact_matrix("Australia","general")
p_age_oz <- import_age_distribution("Australia")
S <- p_age_oz*4e6
E <- rep(10,length(S))
I <- rep(0,length(S))
R <- rep(0,length(S))
```

We first consider the case without any intervention.

```{r, fig.width=7, fig.height=3.5}
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm_oz,dist=p_age_oz)
state0 <- seir_c_state0(S = S,E = E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="No intervention")
```

## Interventions

We can add in time varying interventions using `contact_intervention`. The below example reduces all 
contact by 20%. Simulates physical distancing.

```{r, fig.width=7, fig.height=3.5}
phys_dist = contact_intervention(start = 10,stop = 150,reduce = 0.8,start_delay = 5,stop_delay = 5)
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm_oz,dist=p_age_oz,
                     contact_intervention = phys_dist)
state0 <- seir_c_state0(S = S,E = E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="Physical distancing")
```

the `covoid` package allows you to separate the contact rates into different settings - e.g. school, work and home. 

```{r,fig.height=2.5,fig.width=7}
cm_oz_all <- import_contact_matrix("Australia","general")
cm_oz_sch <- import_contact_matrix("Australia","school")
# separate out school and general population contact rates
cm_oz_all = cm_oz_all - cm_oz_sch
p_all <- plot(cm_oz_all) + labs(title = "General") +
  theme(axis.text.x = element_text(size=6, angle=0),
          axis.text.y = element_text(size=6))
p_sch <- plot(cm_oz_sch) + labs(title = "School") +
  theme(axis.text.x = element_text(size=6, angle=0),
          axis.text.y = element_text(size=6))
gridExtra::grid.arrange(p_all,p_sch,ncol=2)
```

We can then add in time varying interventions using `create_intervention` to each setting. The below example reduces general contact by 20% to simulate physical distancing combined with a reduction in school age contact by 80% to simulate school closures.

```{r, fig.width=7, fig.height=3.5}
cm = list(all = cm_oz_all, sch = cm_oz_sch)
int = list(sch=contact_intervention(start = 10,stop = 150,reduce = 0.2,start_delay = 5,stop_delay = 5),
           all=contact_intervention(start = 10,stop = 150,reduce = 0.8,start_delay = 5,stop_delay = 5))
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm,
                      dist=p_age_oz,contact_intervention = int)
state0 <- seir_c_state0(S = S,E =E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="Physical distancing + school closures")
```

Children under 15 less infectious (as susceptible) and everyone wearing facemasks, hand washing.

```{r, fig.width=7, fig.height=3.5}
im = matrix(1,ncol=16,nrow=16)
im[,1:3] = 0.8

cm = list(all = cm_oz_all, sch = cm_oz_sch)
int_t = transmission_intervention(start = 10,stop = 200,reduce = 0.9,
                                  start_delay = 5,stop_delay = 5)

int_c = list(sch=contact_intervention(start = 10,stop = 150,reduce = 0.2,
                                      start_delay = 5,stop_delay = 5),
           all=contact_intervention(start = 10,stop = 150,reduce = 0.8,
                                    start_delay = 5,stop_delay = 5))
param <- seir_c_param(R0 = 2.5,gamma = 0.1,sigma=0.1,cm=cm,dist=p_age_oz,
                     contact_intervention = int_c,
                     transmission_intervention = int_t,im = im)
state0 <- seir_c_state0(S = S,E =E,I = I,R = R)
res <- simulate_seir_c(t = 500,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="Physical distancing; school closures;\nhandwashing/facemasks; and children less infectious")
```
