---
title: "Models available in COVOID"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.6em;
}
</style>

```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(covoid)
```


## Homogeneous mixing

These models assume....

### SIR model

`sir_state0` `sir_param` and `simulate_sir`

```{r echo=FALSE,fig.cap="*Figure 1a:* SIR model forward dynamics",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SIR {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I[label='I\ninfected'];
  R[label='R\nremoved'];

  edge [color = 'black']
  S -> I
  I -> R
}",height=50)
```

```{r echo=FALSE,fig.cap="*Figure 1b:* SIR model infection dynamics",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SIR {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I[label='I\ninfected'];
  R[label='R\nremoved'];

  edge [color = 'black']
  S -> I
  I -> R
}",height=50)
```

```{r,fig.height=3,fig.width=7}
# A minimal example
param <- sir_param(R0 = 2.5,gamma = 0.1)
state0 <- sir_state0(S = 100,I = 1,R = 0)
res <- simulate_sir(t = 100,state_t0 = state0,param = param)
plot(res,y=c("S","I","R"),main="SIR model")
```

### SEIR model

`seir_state0` `seir_param` and `simulate_seir`

```{r echo=FALSE,fig.cap="*Figure 2a:* SEIR model forward dynamics",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SEIR {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  E[label='E\nexposed\n(latent period)']
  I[label='I\ninfected'];
  R[label='R\nremoved'];

  edge [color = 'black']
  S -> E
  E -> I
  I -> R
}",height=50)
```

```{r  echo=FALSE,fig.cap="*Figure 2b:* SEIR model infection dynamics",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SEIR_inf {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I[label='I\ninfected'];

  edge [color = 'red' style=dashed]
  I -> S
}",height=50)
```


```{r,fig.height=3,fig.width=7}
# A minimal example
param <- seir_param(R0 = 2.5,gamma = 0.1,sigma=0.1)
state0 <- seir_state0(S = 100,E = 1, I = 0,R = 0)
res <- simulate_seir(t = 200,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="SEIR model")
```


### Expanded SEIR model

This model is recommended for modelling COVID-19 (e.g. REFS)

### Expanded SEIR model (Moss et al, 2020)

## Age structured mixing

These models structure the population ... Contact is assumed to be age and setting dependent (see figure). 

```{r echo=FALSE, fig.align="left", fig.height=5, fig.cap= "Figure Y:* Contact matrices for school, work, home and other locations (source: Prem et al ()). Note the differing scales.",fig.width=7}
library(ggplot2)
cm_oz_s <- import_contact_matrix("Australia","school")
cm_oz_w <- import_contact_matrix("Australia","work")
cm_oz_o <- import_contact_matrix("Australia","other")
cm_oz_h <- import_contact_matrix("Australia","home")
p1 = plot(cm_oz_s) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="School",x="Age of individual",y="Age of contact") +
    theme(axis.text = element_text(size = 5))
p2 = plot(cm_oz_w) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="Work",x="Age of individual",y="Age of contact")+
    theme(axis.text = element_text(size = 5))
p4 = plot(cm_oz_o) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="Other",x="Age of individual",y="Age of contact")+
    theme(axis.text = element_text(size = 5))
p3 = plot(cm_oz_h) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="Home",x="Age of individual",y="Age of contact")+
    theme(axis.text = element_text(size = 5))
gridExtra::grid.arrange(p1,p2,p3,p4,ncol=2)
```

### SIR model

The key functions are `sir_c_state0` `sir_c_param` and `simulate_sir_c`

```{r echo=FALSE,fig.cap="*Figure Xa:* Age structured SIR model forward dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SIR_C {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<SUB>j</SUB><BR/><BR/>susceptible>];
  I[label=<I<SUB>j</SUB><BR/><BR/>infected>];
  R[label=<R<SUB>j</SUB><BR/><BR/>removed>];
 
  edge [color = 'black']
  S -> I
  I -> R
}",height=50)
```

```{r echo=FALSE,fig.cap="*Figure Xb:* Age structured SIR model infection dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SEIR_C_inf {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I1[label='I1\ninfected'];
  
  node [color = 'white']
  I2[label='...'];
  
  node [shape = box,
        fontname = Helvetica,
        color='grey']
  IN[label='IN\ninfected'];


  edge [color = 'red' style=dashed]
  I1 -> S
  I2 -> S
  IN -> S
  
}",height=150)
```


```{r,fig.height=3,fig.width=7}
# A minimal example
cm_oz <- import_contact_matrix("Australia","general")
dist_oz <- import_age_distribution("Australia")
param <- sir_c_param(R0 = 2.5,gamma = 0.1,cm=cm_oz,dist=dist_oz)
nJ <- ncol(cm_oz)
S <- rep(100,nJ)
I = rep(1,nJ)
R = rep(0,nJ)
state0 <- sir_c_state0(S = S,I = I,R = R)
res <- simulate_sir_c(t = 150,state_t0 = state0,param = param)
plot(res,y=c("S","I","R"),main="Age structured SIR model")
```

### SEIR model

The key functions for this model are `seir_c_state0` `seir_c_param` and `simulate_seir_c`

```{r echo=FALSE,fig.cap="*Figure Xa:* Age structured SEIR model forward dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SEIR_C {

  graph [overlap = false,
        fontsize = 10,
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<SUB>j</SUB><BR/><BR/>susceptible>];
  E[label=<E<SUB>j</SUB><BR/><BR/>exposed<BR/>(latent period)>];
  I[label=<I<SUB>j</SUB><BR/><BR/>infected>];
  R[label=<R<SUB>j</SUB><BR/><BR/>removed>];
 
  edge [color = 'black']
  S -> E
  E -> I
  I -> R
}",height=50)
```


```{r echo=FALSE,fig.cap="*Figure Xb:* Age structured SEIR model infection dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left"}
DiagrammeR::grViz("
digraph SEIR_C_inf {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<SUB>j</SUB><BR/><BR/>susceptible>];
  I1[label=<I<SUB>1</SUB><BR/><BR/>infected>];
  
  node [color = 'white']
  I2[label='...'];
  
  node [shape = box,
        fontname = Helvetica,
        color='grey']
  IN[label=<I<SUB>J</SUB><BR/><BR/>infected>];

  edge [color = 'red' style=dashed]
  I1 -> S
  I2 -> S
  IN -> S
  
}",height=150)
```

```{r,fig.height=3,fig.width=7}
# A minimal example
cm_oz <- import_contact_matrix("Australia","general")
nJ <- ncol(cm_oz)
dist_oz <- import_age_distribution("Australia")
S <- rep(1000,nJ)
E <- rep(1,nJ)
I <- rep(0,nJ)
R <- rep(0,nJ)
state0 <- seir_c_state0(S = S,E = E,I = I,R = R)
param1 <- seir_c_param(R0 = 2.5,sigma=0.1,gamma = 0.1,cm=cm_oz,dist=dist_oz)
res1 <- simulate_seir_c(t = 200,state_t0 = state0,param = param1)
plot(res1,y=c("S","E","I","R"),main="Age structured SEIR model")
```

## References and notes

All graphs created using [**DiagrammR**](https://rich-iannone.github.io/DiagrammeR/)
