---
title: "Models available in COVOID"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

<style>
p.caption {
  font-size: 0.6em;
}
</style>

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, 
  message = FALSE
)
```

## Overview

```{r setup}
library(covoid)
```

Each model is associated with an initial state, parameter and simulation function. If you are using a model with the code * we pass the parameters for the ODE system to the `*_param` function, the initial compartment states to `*_state0` and finally run the system using `simulate_*`. Models named `*_c` are age-structured, and require a contact matrix and age distribution. These are also the most mature models, all models come with a lifecycle badge making clear the potential for future changes.

## Homogeneous mixing

These models assume the contact rates between individuals in the population can be adequately described by a single value, rather than varying by age or setting.

### SIR model

The key functions for this model are `sir_state0` `sir_param` and `simulate_sir`.

```{r echo=FALSE,fig.cap="*Figure 1a:* SIR model forward dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SIR {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I[label='I\ninfected'];
  R[label='R\nremoved'];

  edge [color = 'black']
  S -> I
  I -> R
}",height=50)
```

```{r echo=FALSE,fig.cap="*Figure 1b:* SIR model infection dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR_inf {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I[label='I\ninfected'];

  edge [color = 'red' style=dashed]
  I -> S
}",height=50)
```

```{r,fig.height=3,fig.width=7,fig.retina=3}
# A minimal example
param <- sir_param(R0 = 2.5,gamma = 0.1)
state0 <- sir_state0(S = 100,I = 1,R = 0)
res <- simulate_sir(t = 100,state_t0 = state0,param = param)
plot(res,y=c("S","I","R"),main="SIR model")
```

### SEIR model

The key functions for this model are `seir_state0` `seir_param` and `simulate_seir`.

```{r echo=FALSE,fig.cap="*Figure 2a:* SEIR model forward dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  E[label='E\nexposed\n(latent period)']
  I[label='I\ninfected'];
  R[label='R\nremoved'];

  edge [color = 'black']
  S -> E
  E -> I
  I -> R
}",height=50)
```

```{r  echo=FALSE,fig.cap="*Figure 2b:* SEIR model infection dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR_inf {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I[label='I\ninfected'];

  edge [color = 'red' style=dashed]
  I -> S
}",height=50)
```


```{r,fig.height=3,fig.width=7,fig.retina=3}
# A minimal example
param <- seir_param(R0 = 2.5,gamma = 0.1,sigma=0.1)
state0 <- seir_state0(S = 100,E = 1, I = 0,R = 0)
res <- simulate_seir(t = 200,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","R"),main="SEIR model")
```


### Expanded SEIR model I

This model is recommended for modelling COVID-19, it is based on @churches2020covoid and @moss2020modelling.

The key functions for this model are `seir1_state0` `seir1_param` and `simulate_seir1`.

```{r echo=FALSE,fig.cap="*Figure 3a:* Expanded SEIR model I forward dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR1 {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']


  S[label=<S<BR/>susceptible>]
  E1[label=<E<SUB>1</SUB><BR/><BR/>latent period<BR/>(asymptomatic)>]
  E2[label=<E<SUB>2</SUB><BR/><BR/>latent period<BR/>(symptomatic)>]
  I1[label=<I<SUB>1</SUB><BR/><BR/>infectious period>];
  I2[label=<I<SUB>2</SUB><BR/><BR/>infectious period<BR/>(mild)>];
  H[label=<H<BR/>hospitalised>];
  R[label=<R<BR/>recovered>];
  F[label=<R<BR/>fatality>];
  E1q[label=<E<SUB>1</SUB><SUP>q</SUP><BR/><BR/>exposed<BR/>(asymptomatic; isolated)>]
  E2q[label=<E<SUB>2</SUB><SUP>q</SUP><BR/><BR/>exposed<BR/>(symptomatic; isolated)>]
  I1q[label=<I<SUB>1</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(isolated)>];
  I2q[label=<I<SUB>2</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(mild; isolated)>];
  Hq[label=<H<SUP>q</SUP><BR/>hospitalised<BR/>(isolated)>];
  Rq[label=<R<BR/>recovered<BR/>(isolated)>];
  Fq[label=<R<BR/>fatality<BR/>(isolated)>];

  edge [color = 'black']
  S -> E1
  E1 -> E2
  E2 -> I1
  I1 -> I2
  I1 -> H
  I2 -> R
  H -> R
  I2 -> F
  H -> F
  S -> E1q
  E1q -> E2q
  E2q -> I1q
  I1q -> I2q
  I1q -> Hq
  I2q -> Rq
  Hq -> Rq
  I2q -> Fq
  Hq -> Fq

}",height=200,width=600)
```

```{r  echo=FALSE,fig.cap="*Figure 3b:* Expanded SEIR model I infection dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR1 {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<BR/>susceptible>]
  E2[label=<E<SUB>2</SUB><BR/><BR/>latent period<BR/>(symptomatic)>]
  I1[label=<I<SUB>1</SUB><BR/><BR/>infectious period>];
  I2[label=<I<SUB>2</SUB><BR/><BR/>infectious period<BR/>(mild)>];
  H[label=<H<BR/>hospitalised>];
  E2q[label=<E<SUB>2</SUB><SUP>q</SUP><BR/><BR/>exposed<BR/>(symptomatic; isolated)>]
  I1q[label=<I<SUB>1</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(isolated)>];
  I2q[label=<I<SUB>1</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(mild; isolated)>];
  Hq[label=<H<SUP>q</SUP><BR/>hospitalised<BR/>(isolated)>];

  edge [color = 'red' style=dashed]
  E2 -> S
  I1 -> S
  I2 -> S
  H -> S
  E2q -> S
  I1q -> S
  I2q -> S
  Hq -> S
}",height=300,width=600)
```

```{r,fig.height=3,fig.width=7,fig.retina=3}
# A minimal example
state0 <- seir1_state0(S = 1e5, E1 = 90, E2 = 40)
param <- seir1_param(R0=2.5,sigma1=0.2,sigma2=0.2,gamma1=0.2,gamma2=0.2,gamma3=0.2,
                     Qeff=0.5,Heff=0.9,rho=0.1,alpha=0.1,eta=0.02)
res <- simulate_seir1(t = 250,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","Recov","Fatal"),main="Expanded SEIR model I")
```

### Expanded SEIR model II

This model is recommended for modelling COVID-19, it is a (partial) reproduction of @moss2020modelling.

The key functions for this model are `seir2_state0` `seir2_param` and `simulate_seir2`.

```{r echo=FALSE,fig.cap="*Figure 4a:* Expanded SEIR model II (Doherty model) forward dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR2 {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']


  S[label=<S<BR/>susceptible>]
  E1[label=<E<SUB>1</SUB><BR/><BR/>latent period<BR/>(asymptomatic)>]
  E2[label=<E<SUB>2</SUB><BR/><BR/>latent period<BR/>(symptomatic)>]
  I1[label=<I<SUB>1</SUB><BR/><BR/>infectious period>];
  I2[label=<I<SUB>2</SUB><BR/><BR/>infectious period<BR/>(mild)>];
  M[label=<H<BR/>medical management>];
  R[label=<R<BR/>recovered>];
  E1q[label=<E<SUB>1</SUB><SUP>q</SUP><BR/><BR/>exposed<BR/>(asymptomatic; isolated)>]
  E2q[label=<E<SUB>2</SUB><SUP>q</SUP><BR/><BR/>exposed<BR/>(symptomatic; isolated)>]
  I1q[label=<I<SUB>1</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(isolated)>];
  I2q[label=<I<SUB>2</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(mild; isolated)>];
  Mq[label=<H<SUP>q</SUP><BR/>medical management<BR/>(isolated)>];
  Rq[label=<R<BR/>recovered<BR/>(isolated)>];
  CTnm[label=<CT<SUB>nm</SUB><BR/><BR/>Contact of<BR/>non-managed case>];
  CTm[label=<CT<SUB>nm</SUB><BR/><BR/>Contact of<BR/>managed case>];

  edge [color = 'black']
  S -> E1
  E1 -> E2
  E2 -> I1
  I1 -> I2
  I1 -> M
  I2 -> R
  M -> R
  S -> E1q
  E1q -> E2q
  E2q -> I1q
  I1q -> I2q
  I1q -> Hq
  I2q -> Rq
  Mq -> Rq

}",height=200,width=600)
```

```{r  echo=FALSE,fig.cap="*Figure 4b:* Expanded SEIR model II (Doherty model) infection dynamics",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR2 {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<BR/>susceptible>]
  E2[label=<E<SUB>2</SUB><BR/><BR/>latent period<BR/>(symptomatic)>]
  I1[label=<I<SUB>1</SUB><BR/><BR/>infectious period>];
  I2[label=<I<SUB>2</SUB><BR/><BR/>infectious period<BR/>(mild)>];
  M[label=<H<BR/>medical management>];
  E2q[label=<E<SUB>2</SUB><SUP>q</SUP><BR/><BR/>exposed<BR/>(symptomatic; isolated)>]
  I1q[label=<I<SUB>1</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(isolated)>];
  I2q[label=<I<SUB>1</SUB><SUP>q</SUP><BR/><BR/>infectious period<BR/>(mild; isolated)>];
  Mq[label=<H<SUP>q</SUP><BR/>medical management<BR/>(isolated)>];

  edge [color = 'red' style=dashed]
  E2 -> S
  I1 -> S
  I2 -> S
  M -> S
  E2q -> S
  I1q -> S
  I2q -> S
  Mq -> S
}",height=300,width=600)
```

```{r,fig.height=3,fig.width=7,fig.retina=3}
# A minimal example
state0 <- seir2_state0(S = 1e5, E1 = 90, E2 = 40)
param <- seir2_param(R0=2.5,lambdaimp=0,sigma1=0.2,sigma2=0.2,gamma1=0.2,gamma2=0.2,
                    gammaq1=0.1,gammaq2=0.1,Qeff=0.5,Meff=0.99,rho=0.0,eta=1/sqrt(2),
                      alphamBeta=0.5,probHospGivenInf=0.09895,delta=1/14,kappa=20,pm=1)
res <- simulate_seir2(t = 250,state_t0 = state0,param = param)
plot(res,y=c("S","E","I","Recov","Fatal"),main="Moss et al (2020) model")
```

## Age structured mixing

These models structure the population by discrete age partitions, with the contact rate assumed to be age and setting dependent (see figure). 

```{r echo=FALSE, fig.align="left", fig.height=5, fig.cap= "Figure 5:* Contact matrices for school, work, home and other locations (source: Prem, Cook and Jit (2017)). Note the differing scales.",fig.width=7,fig.retina=3}
library(ggplot2)
cm_oz_s <- import_contact_matrix("Australia","school")
cm_oz_w <- import_contact_matrix("Australia","work")
cm_oz_o <- import_contact_matrix("Australia","other")
cm_oz_h <- import_contact_matrix("Australia","home")
p1 = plot(cm_oz_s) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="School",x="Age of individual",y="Age of contact") +
    theme(axis.text = element_text(size = 5))
p2 = plot(cm_oz_w) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="Work",x="Age of individual",y="Age of contact")+
    theme(axis.text = element_text(size = 5))
p4 = plot(cm_oz_o) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="Other",x="Age of individual",y="Age of contact")+
    theme(axis.text = element_text(size = 5))
p3 = plot(cm_oz_h) + 
    scale_fill_continuous(name="Contact\nrate",type = "viridis") +
    labs(title="Home",x="Age of individual",y="Age of contact")+
    theme(axis.text = element_text(size = 5))
gridExtra::grid.arrange(p1,p2,p3,p4,ncol=2)
```

### SIR model

The key functions for this model are `sir_c_state0` `sir_c_param` and `simulate_sir_c`.

```{r echo=FALSE,fig.cap="*Figure 6a:* Age structured SIR model forward dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SIR_C {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<SUB>j</SUB><BR/><BR/>susceptible>];
  I[label=<I<SUB>j</SUB><BR/><BR/>infected>];
  R[label=<R<SUB>j</SUB><BR/><BR/>removed>];
 
  edge [color = 'black']
  S -> I
  I -> R
}",height=50)
```

```{r echo=FALSE,fig.cap="*Figure 6b:* Age structured SIR model infection dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR_C_inf {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label='S\nsusceptible'];
  I1[label=<I<SUB>1</SUB><BR/><BR/>infectious>];
  
  node [color = 'white']
  I2[label='...'];
  
  node [shape = box,
        fontname = Helvetica,
        color='grey']
  IN[label=<I<SUB>n</SUB><BR/><BR/>infectious>];


  edge [color = 'red' style=dashed]
  I1 -> S
  I2 -> S
  IN -> S
  
}",height=150)
```

```{r,fig.height=3,fig.width=7,fig.retina=3}
# A minimal example
cm_oz <- import_contact_matrix("Australia","general")
dist_oz <- import_age_distribution("Australia")
param <- sir_c_param(R0 = 2.5,gamma = 0.1,cm=cm_oz,dist=dist_oz)
nJ <- ncol(cm_oz)
S <- rep(100,nJ)
I <- rep(1,nJ)
R <- rep(0,nJ)
state0 <- sir_c_state0(S = S,I = I,R = R)
res <- simulate_sir_c(t = 150,state_t0 = state0,param = param)
plot(res,y=c("S","I","R"),main="Age structured SIR model")
```

### SEIR model

The key functions for this model are `seir_c_state0` `seir_c_param` and `simulate_seir_c`.

```{r echo=FALSE,fig.cap="*Figure 7a:* Age structured SEIR model forward dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR_C {

  graph [overlap = false,
        fontsize = 10,
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<SUB>j</SUB><BR/><BR/>susceptible>];
  E[label=<E<SUB>j</SUB><BR/><BR/>exposed<BR/>(latent period)>];
  I[label=<I<SUB>j</SUB><BR/><BR/>infected>];
  R[label=<R<SUB>j</SUB><BR/><BR/>removed>];
 
  edge [color = 'black']
  S -> E
  E -> I
  I -> R
}",height=50)
```


```{r echo=FALSE,fig.cap="*Figure 7b:* Age structured SEIR model infection dynamics. There are j=1,...,J age groups.",fig.height=200,fig.align="left",fig.retina=3}
DiagrammeR::grViz("
digraph SEIR_C_inf {

  graph [overlap = false,
        fontsize = 10
        rankdir = LR]

  node [shape = box,
        fontname = Helvetica,
        color='grey']

  S[label=<S<SUB>j</SUB><BR/><BR/>susceptible>];
  I1[label=<I<SUB>1</SUB><BR/><BR/>infected>];
  
  node [color = 'white']
  I2[label='...'];
  
  node [shape = box,
        fontname = Helvetica,
        color='grey']
  IN[label=<I<SUB>J</SUB><BR/><BR/>infected>];

  edge [color = 'red' style=dashed]
  I1 -> S
  I2 -> S
  IN -> S
  
}",height=150)
```

```{r,fig.height=3,fig.width=7,fig.retina=3}
# A minimal example
cm_oz <- import_contact_matrix("Australia","general")
nJ <- ncol(cm_oz)
dist_oz <- import_age_distribution("Australia")
S <- rep(1000,nJ)
E <- rep(1,nJ)
I <- rep(0,nJ)
R <- rep(0,nJ)
state0 <- seir_c_state0(S = S,E = E,I = I,R = R)
param1 <- seir_c_param(R0 = 2.5,sigma=0.1,gamma = 0.1,cm=cm_oz,dist=dist_oz)
res1 <- simulate_seir_c(t = 200,state_t0 = state0,param = param1)
plot(res1,y=c("S","E","I","R"),main="Age structured SEIR model")
```

## References and notes

All graphs created using [**DiagrammR**](https://rich-iannone.github.io/DiagrammeR/)


